SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2014-03-14 paper/paper.tex
grammar.
paper: tweak, commit .pdf so folks can read it.

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is extrapolation from
individual counterexamples to a class of counterexamples.  This leap of
abstraction is often implicitly made by the programmer in determining the reason
why counterexamples fail the property.  For example, the following is a
relatively small counterexample returned when QuickChecking a property in (a
buggy version of) Xmonad, a popular X11 window manager written in
Haskell~\cite{xmonad}:
%
\medskip%
\begin{sloppypar}
%\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
(This counterexample uses Haskell's default \ttp{Show} instances, which uses
record syntax.)  Functional programmers should be familiar with having to debug
a ``wall of text'' when debugging data structures.  What if instead a
formula like the following were returned, stating that for any well-typed
values $x_0$, $x_1$, $x_2$, and $x_3$, tested, a counterexample is found?

@@

\paragraph{SmartCheck}
Motivated by the problems of reducing and generalizing large counterexamples, we
developed \emph{SmartCheck}.  SmartCheck takes a counterexample produced by some
oracle (we use QuickCheck) and generically minimizes and generalizes the
counterexample.

As demonstrated by a motivating example in Section~\ref{sec:example}, SmartCheck
can shrink more than an order-of-magnitude faster than efficient hand-written
shrink implementations for some properties and programs.  As well, the time
required to shrink scales linearly with the size of the counterexample---other
common shrink implementations often scale in polynomial time or worse.
(Shrinking time is important in real-world uses; see
Section~\ref{sec:experiments} for examples.)  And the counterexamples returned
are two to five times smaller.

@@

SmartCheck implements three novel approaches to automatically generalize
counterexamples, which are described in Section~\ref{sec:generalization}.  The
first algorithm universally quantifies sub-values that always fail in tests, as
briefly described earlier in the introduction.  The second algorithm
existentially quantifies sub-values for types in which every possible variant
fails the property.  For example, finding counterexamples (\ttp{Left 2}) and
(\ttp{Right True}) for the type
%
\begin{code}
\ttp{Either Int Bool}
\end{code}
%
\noindent
means there exists a counterexample regardless of the variant chosen.
Existential generalization is important for large sum types, as found in
abstract syntax tree (AST) definitions, for example.

The third algorithm automatically strengthens properties by omitting ``similar''
counterexamples to the ones previously observed.  The algorithm is motivated by
noting that there are often multiple ways in which a property may fail; for
example, a property that parsing a pretty-printed AST results in the original
AST may fail due to multiple bugs.  During testing, it is useful to discover
counterexamples arising from each bug in one go.  In practice, the problem is
solved by discovering a counterexample \ttp{cex}, abstracting it, and then
adding a new precondition to the property that informally says ``omit
counterexamples of form \ttp{cex}.''  Adding preconditions manually is laborious
and may cause the programmer to make premature fixes to the program, if she
believes she has isolated the error before she actually does.

@@

\begin{figure}[ht]
\begin{code}
data N a = N a
type I   = [N Int16]
data T   = T (N Word8) I I I I

toList :: T -> [[Int16]]
toList (T w i0 i1 i2 i3) =
  [go w] : (map . map) go [i0, i1, i2, i3]
  where go (N i) = fromIntegral i

@@

In this section, we motivate in more detail the challenges in shrinking
counterexamples (we focus on shrinking and only touch on generalization here).
Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.  We use Haskell \ttp{Prelude} functions, \ttp{mapMaybe} from
  \ttp{Data.Maybe}, and \ttp{liftM} and \ttp{replicateM} from
  \ttp{Control.Monad}.}  Data type \ttp{T} is a product type containing four
lists of wrapped \ttp{Int16}s and a single wrapped \ttp{Word8}.  For our
purposes, it is not important what \ttp{T} models, but for concreteness, suppose
it models the values in four linked lists in an operating system, together with
an 8-bit status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
satisfies \ttp{pre} but fails \ttp{post} (the \ttp{==>}) operator in the figure
is implication from the QuickCheck library).

Despite the simplicity of the example, a typical counterexample returned by
QuickCheck can be large.  With standard settings and no shrinking, the median
counterexample discovered contains around 70 \ttp{N} constructors, and
counterexamples can contain over 100 \ttp{N} constructors.  Thus, it pays to define
\ttp{shrink}!

@@

Still, using this definition, from some initial counterexamples, shrinking may
take over one minute and result in a final counterexample containing just under
100 \ttp{N} constructors.\footnote{All results reported in the paper are with a
  GHC-7.6.1-compiled program, using \ttp{-O2}, on a 4-core 2.7GHz Intel~i7
  processor.}

@@

In regards to generalization, after shrinking the counterexample, SmartCheck
will also generalize it.  Note that for any \ttp{Word8} value in the first
argument of \ttp{T}, the counterexample still holds.  For example, SmartCheck
might return a counterexample of the form
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}

@@

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
Recall from the end of Section~\ref{sec:example} the quantified formula
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
characterizing a set of counterexamples.  Another example using the calculator
language from Section~\ref{sec:subval} is to generalize a counterexample like
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
\noindent
by
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
since any dividend results in divide-by-zero for the given divisor.  Not only do
the generalizations assist the programmer's insight, but they reduce the sheer
size of the counterexample.  We call the kind of formula just shown
\emph{universal sub-value generalization} and it is implemented in SmartCheck.

@@

The data type to be tested must derive the \ttp{Read}, \ttp{Show},
\ttp{Typeable}, and \ttp{Generic} type classes.  (We only require \ttp{Read}
because standard QuickCheck does not return a counterexample; we read the shown
values.)  Deriving \ttp{Typeable} and \ttp{Generic} require using the GHC
language extensions \ttp{DeriveDataTypeable} and \ttp{DeriveGeneric},
respectively.  \ttp{Typeable} is used for dynamic typing in defining the
\ttp{replace} method of \ttp{SubTypes} since it is unknown at compile-time
whether the value to be replaced and replacing value have the same types.
However, through SmartCheck's interface, run-time failures due to
type-mismatches will not occur.

@@

A property may be a function taking any number of arguments, but SmartCheck only
operates on a single input during testing, which is assumed to be the first
argument to the function.  During all stages of SmartCheck's execution, the
remaining arguments are held constant.  For example, for a property of type
%
\begin{code}
prop :: Foo -> Int -> Bool
\end{code}
%
\noindent
and a pair of arguments for which \ttp{prop} returns \ttp{False}
%
\begin{code}
(SomeFoo ...) 3
\end{code}
%
\noindent
returned by QuickCheck, SmartCheck reduces and generalizes \ttp{SomeFoo} with
respect to the property
%
\begin{code}
prop' :: Foo -> Bool
prop' foo = prop foo 3
\end{code}

Our experience is that often a program operates over one large data structure
together with smaller inputs.  While multiple inputs could be analyzed
simultaneously by SmartCheck, it is not obvious how to clearly represent
multiple large counterexamples to the user---the implementation currently prints
to standard out.  (In future work, a web-based interface could help.)

@@

The interface types and functions for SmartCheck with analogous behavior to
QuickCheck's are prefixed with an \ttp{sc} to avoid name space collisions with
QuickCheck.  So the return type of property must be either \ttp{Bool} or
\ttp{ScProperty}, a data type that corresponds to QuickCheck's \ttp{Property}
type.  A function that returns \ttp{Bool} can by turned into one that returns
\ttp{ScProperty} using the \ttp{scProperty} function (corresponding to
QuickCheck's \ttp{property} function).  SmartCheck's property implication uses
an arrow \ttp{-->} corresponding to QuickCheck's \ttp{==>}.

The arguments is a record type that includes the values described above to
control SmartCheck's behavior.  The standard arguments are contained in the
value \ttp{scStdArgs}.

@@

\section{Experiments}\label{sec:experiments}
Here we describe two experiments using SmartCheck, including an XMonad
property and a property about a natural language processing library.

@@

\subsection{XMonad}
Recall from the introduction the XMonad example.  The XMonad window manager is a
large software project with many contributors, so consequently, a QuickCheck
test harness is included to help ensure new commits do not introduce bugs.  At
the heart of XMonad is a \ttp{StackSet} data type that encodes the relationship
between windows, work spaces, and which window has the focus.  XMonad contains
properties to ensure the correct manipulation of \ttp{StackSet}s.

XMonad passes all of its QuickCheck tests, but let us see what might happen to a
new contributor if things go awry.  Suppose a developer defines a deletion
function to delete a window, if it exists.  There is a deletion function that
exists in XMonad, which is quite complex, given the amount of state that is managed
by \ttp{StackSet}.  However, one function used in deletion is to filter the
stack of windows associated with each workspace defined:
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (/=w) \}
\end{code}
%
\noindent
Now, suppose the programmer makes a simple typo and instead writes
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (==w) \}
\end{code}
%
\noindent
When testing the property \ttp{prop\_delete}, which says that deleting the
focused window of the current stack removes it from the \ttp{StackSet} \ttp{x}.
%
\begin{code}
prop_delete x =
    case peek x of
        Nothing -> True
        Just i  -> not (member i (delete i x))
\end{code}
%
\noindent
QuickCheck returns the large value shown in the introduction.  That value is a
relatively small counterexample, but even the smallest \ttp{StackSet} values
are somewhat large due to the number of fields within it.  Recall the value
returned by SmartCheck after generalization:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
Let us examine what was generalized and what was not.  First, the last element
of \ttp{StackSet} is a map, and we treated it as a opaque type, so it is not
analyzed at all.  But it is an empty map.  The second and third fields of
\ttp{StackSet} are generalized, but the first one is not.  So there is something
particular about it.  So the culprit is one of the small constants (1 and -1) or
having a \ttp{Just} value rather than a \ttp{Nothing}: it turns out that what
matters is having a \ttp{Just} value, which is the stack field that deletion
works on.

@@

\subsection{Natural Language Processing}
In 2012, a question was posted on the programming message board
Stack Overflow asking about how to shrink large nested data
types.\footnote{\url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}}
The poster writes:
\begin{quote}
$\ldots$ I tend to get an incomprehensible page full of output. $\ldots$
  Implementing the shrink function for each of my types seems to help a little,
  but not as much as I'd like.  $\ldots$ If I try to tune my shrink
  implementations, I also find that QC starts taking a very long time.
\end{quote}
\noindent
The question relates to the Geni natural language processing (NLP) package
implemented in Haskell~\cite{kow}.  Specifically, counterexamples to a property
attempting to show that a macro expansion function is its own inverse are
enormous, requiring 200-300 80-character lines to print.

@@

We were able to send a substantially reduced and generalized counterexample to
the message poster, making the cause of the bug more obvious.  He responded:
``$\ldots$While your improved shrinking may not have gone `all'
the way to the bottom, it got me a \emph{huge} chunk of the way there!''  Recall
that through the entire process, we never had to learn how GenI works, what the
property meant, or how to write a custom shrink method!

``Retro-fitting'' SmartCheck to GenI, which is a large software system, required
some work.  Fortunately, because the program already uses QuickCheck,
\ttp{arbitrary} instances were already defined, but we had to hunt down and
define default instances for a large number of definitions.

@@

We have benchmarked Feat as well on the example in the introduction
(Figure~\ref{fig:initial}).  It performs quite competitively to SmartCheck
in terms the size of counterexamples discovered: the mean and median number of
\ttp{N} constructors in the counterexamples returned by Feat are both seven.
However, Feat performs less well with respect to how long counterexample
discovery takes: the mean time taken by Feat is 5.60 seconds, with a long tail:
five percent of runs take over 14 seconds (the mean time taken by
SmartCheck is 0.12 seconds, and 99\% of runs take less than 0.30 seconds).

Finally, SmartCheck bears some similarity to QuickSpec, a testing-based library
that infers equational properties about programs~\cite{qs} insofar as they both
attempt to generalize counterexamples based on specific inputs.  QuickSpec
attempts to infer equational properties of programs through random testing.
Similarly, Daikon infers assertions for C, C++, Java, and Perl by observing
relationships between variables in executions of a program~\cite{daikon}.
SmartCheck does not attempt to infer properties like these tools do, so it is in
one sense less general.  However, SmartCheck's three kinds counterexample
generalization are novel.

@@

\section*{Acknowledgments}
I thank Andy Gill, Joe Hurd, John Launchbury, Simon Winwood, and especially John
Hughes for reading an earlier draft of this paper.  Their deep insights and bug
catches dramatically improved both the form and content.

>>>>

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is extrapolation from
individual counterexamples to a class of counterexamples.  This leap of
abstraction is often implicitly made by the programmer in determining the reason
why counterexamples fail the property.  For example, the following is a
relatively small counterexample returned when QuickChecking a property in (a
bug-injected version of) Xmonad, a popular X11 window manager written in
Haskell~\cite{xmonad}:
%
\medskip%
\begin{sloppypar}
%\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
(This counterexample uses Haskell's default \ttp{Show} instances, which uses
record syntax.)  Programmers should be familiar with having to debug
a ``wall of text'' as shown above.  What if instead a
formula like the following were returned, stating that for any well-typed
values $x_0$, $x_1$, $x_2$, and $x_3$, tested, a counterexample is found?

@@

\paragraph{SmartCheck}
Motivated by the problems of reducing and generalizing large counterexamples, we
developed \emph{SmartCheck}.  SmartCheck takes a counterexample produced by some
oracle and generically minimizes and generalizes the counterexample.

As shown in the motivating example in Section~\ref{sec:example}, SmartCheck
can shrink more than an order-of-magnitude faster than efficient hand-written
shrink implementations for some properties and programs.  As well, the time
required to shrink scales linearly with the size of the counterexample---other
common shrink implementations often scale in polynomial time or worse.
(Shrinking time is important in real-world uses; see
Section~\ref{sec:experiments} for examples.)

@@

SmartCheck implements three novel approaches to automatically generalize
counterexamples, which are described in Section~\ref{sec:generalization}.  The
first algorithm universally quantifies sub-values that always fail in tests.
The second algorithm existentially quantifies sub-values for types in which
every possible variant fails the property.  For example, finding counterexamples
(\ttp{Left 2}) and (\ttp{Right True}) for the type
%
\begin{code}
\ttp{Either Int Bool}
\end{code}
%
\noindent
means there exists a counterexample regardless of the variant chosen.
Existential generalization is useful for large sum types, as found in
abstract syntax tree (AST) definitions, for example.

The third algorithm automatically strengthens properties by omitting ``similar''
counterexamples to the ones previously observed.  The algorithm is motivated by
noting that there are often multiple ways in which a property may fail; for
example, a property stating that pretty-printing an AST and then parsing it
results in the original AST may fail due to multiple bugs.  During testing, it
is useful to discover counterexamples arising from each bug in one go.  In
practice, the problem is solved by discovering a counterexample \ttp{cex},
abstracting it, and then adding a new precondition to the property that
informally says ``omit counterexamples of form \ttp{cex}.''  Adding
preconditions manually is laborious and may cause the programmer to make
premature fixes to the program, if she believes she has isolated the error
before she actually does.

@@

\begin{figure}[ht]
\begin{code}
type I   = [Int16]
data T   = T (Word8) I I I I

toList :: T -> [[Int16]]
toList (T w i0 i1 i2 i3) =
  [fromIntegral w] : (map . map) fromIntegral rst
  where
  rst = [i0, i1, i2, i3]

@@

\todo{redo analysis, check text}
In this section, we motivate in more detail the challenges in shrinking
counterexamples (we focus on shrinking and only touch on generalization here).
Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.  We use Haskell \ttp{Prelude} functions, \ttp{mapMaybe} from
  \ttp{Data.Maybe}, and \ttp{liftM} and \ttp{replicateM} from
  \ttp{Control.Monad}.}  Data type \ttp{T} is a product type containing four
lists of wrapped \ttp{Int16}s and a single wrapped \ttp{Word8}.  For our
purposes, it is not important what \ttp{T} models, but for concreteness, suppose
it models the values in four linked lists in an operating system, together with
an 8-bit status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T 10 [-20000] [-20000] [] []
\end{code}
%
\noindent
satisfies \ttp{pre} but fails \ttp{post} (the \ttp{==>} operator in the figure
is implication from the QuickCheck library).

Despite the simplicity of the example, a typical counterexample returned by
QuickCheck can be large.  With standard settings and no shrinking, the median
counterexample discovered contains around 70 \ttp{Int16} values, and
counterexamples can contain over 100 values.  \todo{recheck} Thus, it pays to
define \ttp{shrink}!

@@

Still, using this definition, from some initial counterexamples, shrinking may
take over one minute and result in a final counterexample containing just under
100 \ttp{Int16} values.\footnote{All results reported in the paper are with a
  GHC-7.6.3-compiled program, using \ttp{-O2}, on a 4-core 2.7GHz Intel~i7
  processor.}

@@

In regards to generalization, after shrinking the counterexample, SmartCheck
will also generalize it.  Note that for any \ttp{Word8} value in the first
argument of \ttp{T}, the counterexample still holds.  For example, SmartCheck
might return a counterexample of the form
%
\begin{code}
forall x . T x [-20000] [-20000] [] []
\end{code}

@@

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
Recall from the end of Section~\ref{sec:example} the quantified formula
%
\begin{code}
forall x . T x [-20000] [-20000] [] []
\end{code}
%
\noindent
characterizing a set of counterexamples.  Another example using the calculator
language from Section~\ref{sec:subval} is to generalize a counterexample like
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
\noindent
by
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
since any dividend results in divide-by-zero for the given divisor.  Not only do
the generalizations assist the programmer's insight, but they reduce the sheer
size of the counterexample.  We call the kind of formula just shown
\emph{universal sub-value generalization} and it is implemented in SmartCheck.

@@

The data type to be tested must derive the \ttp{Typeable}, and \ttp{Generic}
type classes.  Deriving \ttp{Typeable} and \ttp{Generic} require using the GHC
language extensions \ttp{DeriveDataTypeable} and \ttp{DeriveGeneric},
respectively.  \ttp{Typeable} is used for dynamic typing in defining the
\ttp{replace} method of \ttp{SubTypes} since it is unknown at compile-time
whether the value to be replaced and replacing value have the same types.
However, through SmartCheck's interface, run-time failures due to
type-mismatches will not occur.  Additionally, like with QuickCheck, deriving
\ttp{Show} is required to print counterexamples discovered.

@@

The kinds of programs SmartCheck is specialized for are ones that operate over a
large data structure together with smaller inputs.  Therefore, properties
provided to SmartCheck are expected to be of the form
%
\begin{code}
Testable prop => a -> prop
\end{code}
%
where \ttp{a} is the type of the value for SmartCheck to analyze, and \ttp{prop}
is a testable property, as defined by QuickCheck; morally, these are functions
(or degenerately, values) that evaluate to a Boolean value.

If QuickCheck is used to discover a counterexample, all arguments except the
first are shrunk, if their types have \ttp{shrink} methods defined for them.
The first argument is returned to SmartCheck to be shrunk or generalized
according to the algorithms described earlier.

%% \todo{is still subject to QC's maxsize constraints}
%% During all stages of SmartCheck's execution, the
%% remaining arguments are held constant.  For example, for a property of type
%% %
%% \begin{code}
%% prop :: Foo -> Int -> Bool
%% \end{code}
%% %
%% \noindent
%% and a pair of arguments for which \ttp{prop} returns \ttp{False}
%% %
%% \begin{code}
%% (SomeFoo ...) 3
%% \end{code}
%% %
%% \noindent
%% returned by QuickCheck, SmartCheck reduces and generalizes \ttp{SomeFoo} with
%% respect to the property
%% %
%% \begin{code}
%% prop' :: Foo -> Bool
%% prop' foo = prop foo 3
%% \end{code}

@@

The interface types and functions for SmartCheck with analogous behavior to
QuickCheck's are prefixed with an \ttp{sc} to avoid name space collisions with
QuickCheck.  Others are specialized for SmartCheck; e.g., enabling or disabling
universal or existential extrapolated, number of extrapolation rounds, and
limits on the depth and size of the values to generate.

@@

\section{Experiments}\label{sec:experiments}
We describe two experiments using SmartCheck, including an XMonad property and a
property about a natural language processing library.

@@

\subsection{XMonad}
Recall from the introduction the XMonad example.  The XMonad window manager is a
large software project with many contributors, so naturally, a QuickCheck
test harness is included to help ensure new commits do not introduce bugs.  At
the heart of XMonad is a \ttp{StackSet} data type that encodes the relationship
between windows, work spaces, and which window has the focus.  XMonad contains
properties to ensure the correct manipulation of \ttp{StackSet}s.  Due to having
one large data-structure that is essential to the entire program, XMonad is a
perfect candidate for SmartCheck.

XMonad passes all of its QuickCheck tests, but let us see what might happen to a
new contributor if things go awry.  Suppose a developer defines a deletion
function to delete a window, if it exists.  There is a deletion function that
exists in XMonad, which is quite complex, given the amount of state that is managed
by \ttp{StackSet}.  However, one function used in deletion is to filter the
stack of windows associated with each workspace defined:
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (/=w) \}
\end{code}
%
\noindent
Now, suppose the programmer makes a simple typo and instead writes
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (==w) \}
\end{code}
%
\noindent
When testing the property \ttp{prop\_delete}, which says that deleting the
focused window of the current stack removes it from the \ttp{StackSet} \ttp{x}.
%
\begin{code}
prop_delete x =
    case peek x of
        Nothing -> True
        Just i  -> not (member i (delete i x))
\end{code}
%
\noindent
QuickCheck returns the large value shown in the introduction.  That value is a
relatively small counterexample, but even the smallest \ttp{StackSet} values are
somewhat visually overwhelming due to the number of fields within it.  Recall
the value returned by SmartCheck after generalization:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
Let us examine what was generalized.  In our test run, we chose to treat data
maps as opaque, so the forth element of \ttp{StackSet} is not generalized, but
is simply the empty map, so looks uninteresting.  The second and third fields of
\ttp{StackSet} are generalized, but the first one is not.  So there is something
particular about it.  So the culprit is one of the small constants (1 and -1) or
having a \ttp{Just} value rather than a \ttp{Nothing}: it turns out that what
matters is having a \ttp{Just} value, which is the stack field that deletion
works on!

@@

\subsection{Natural Language Processing}
In 2012, a question was posted on the programming message board
Stack Overflow asking about how to shrink large nested data
types.\footnote{\url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}}
The poster writes:
\begin{quote}
$\ldots$ I tend to get an incomprehensible page full of output. $\ldots$
  Implementing the shrink function for each of my types seems to help a little,
  but not as much as I'd like.  $\ldots$ If I try to tune my shrink
  implementations, I also find that QC starts taking a very long time.
\end{quote}

\todo{how long unanswered?}

\noindent
The question relates to the Geni natural language processing (NLP) package
implemented in Haskell~\cite{kow}.  Specifically, counterexamples to a property
attempting to show that a macro expansion function is its own inverse are
enormous, requiring 200-300 80-character lines to print.

@@

We were able to send a substantially reduced and generalized counterexample to
the message poster, making the cause of the bug more obvious.  The author responded:
\begin{quote}
$\ldots$While your improved shrinking may not have gone `all'
the way to the bottom, it got me a \emph{huge} chunk of the way there!
\end{quote}

\noindent
Recall that through the entire process, we never had to learn how GenI works,
what the property meant, or how to write a custom shrink method!

``Retro-fitting'' SmartCheck to GenI, which is a large software system, required
some boilerplate work.  Fortunately, because the program already uses
QuickCheck, \ttp{arbitrary} instances were already defined, but we had to hunt
down and define default instances for a large number of definitions (indeed, the
job could have largely been done with Sed and Awk).

@@

We have benchmarked Feat as well on the example in the introduction
(Figure~\ref{fig:initial}).  It performs quite competitively to SmartCheck
in terms the size of counterexamples discovered: the mean and median number of
\ttp{N} constructors in the counterexamples returned by Feat are both seven. \todo{check}
However, Feat performs less well with respect to how long counterexample
discovery takes: the mean time taken by Feat is 5.60 seconds, with a long tail:
five percent of runs take over 14 seconds (the mean time taken by
SmartCheck is 0.12 seconds, and 99\% of runs take less than 0.30 seconds).

Finally, SmartCheck bears some similarity to QuickSpec, a testing-based library
that infers equational properties about programs~\cite{qs} insofar as they both
attempt to generalize counterexamples based on specific inputs.  QuickSpec
attempts to infer equational properties of programs through random testing.
Similarly, Daikon infers assertions for C, C++, Java, and Perl by observing
relationships between variables in executions of a program~\cite{daikon}.
SmartCheck does not attempt to infer properties like these tools do, so it is in
one sense less general.  However, SmartCheck's three kinds of counterexample
generalization are novel.

@@

\todo{mention memoization (generalized tries)}

\todo{tail-recursive algorithm is fast and easy to port to imperative languages}

\section*{Acknowledgments}
I thank Andy Gill, Joe Hurd, John Launchbury, Simon Winwood, and especially John
Hughes for reading an earlier draft of this paper.  Their deep insights and bug
catches dramatically improved both the form and content.  This paper was
submitted (and rejected) from ICFP'13; the anonymous reviewers' suggestions
improved the presentation and especially the implementation.
